Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-02-07 LaTeX/S2Appendix.tex
moved appendices out of main text but can crossref

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Haploid selection, sex ratio bias, and transitions between sex-determining systems} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Michael Francis Scott\textsuperscript{1\Yinyang*},
Matthew Miles Osmond\textsuperscript{2\Yinyang},
Sarah Perin Otto\textsuperscript{2}
\\
\bigskip
\textbf{1} UCL Genetics Institute, Department of Genetics, Evolution and Environment, University College London, Gower Street, London WC1E 6BT 
\\
\textbf{2} Department of Zoology, University of British Columbia, \#4200 - 6270 University Boulevard, Vancouver, BC, Canada V6T 1Z4
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* m.f.scott@ucl.ac.uk

\end{flushleft}
% Please keep the abstract below 300 words
%currently 312 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a sexually-antagonistic locus. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or unlinked) to the locus under selection, which is not predicted by previous theory. 
(2) We also consider selection upon haploid genotypes either during gametic competition (e.g., pollen/sperm competition) or meiosis (i.e., non-Mendelian segregation), selective processes that typically occur in one sex or the other. 
We find that associations with haploid selected loci can drive transitions between sex-determining systems, without requiring sexually-antagonistic selection in diploids. 
Unexpectedly, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Furthermore, haploid selection in the heterogametic sex can cause sex ratio biases, which may increase or decrease with the spread of new sex chromosomes. 
Thus, we find that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
These results predict conditions under which sex-determining systems are likely to be labile and draw novel connections with sex ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Accumulating evidence indicates that systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models extends these theories to include selection upon haploid genotypes (meiotic drive or gametic competition), which causes sex ratio biases and occurs differently in males and females. 
Surprisingly, we find that neither force (sex ratio selection nor associations with genes that have sex-specific effects) dominates the spread of new sex-determining systems alone. 
Even more unexpectedly, we find that, to spread, new sex-determining alleles do not necessarily have to arise in closer linkage with genes that are differentially selected in males and females. 
Therefore, our models predict loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}). 
Among species with genetic sex determination (GSD) of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes. 
For example, transitions of the master sex-determining locus between chromosomes or the evolution of new master sex-determining loci where the heterogametic sex does not change (hereafter `cis-GSD transitions') have occurred in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many clades exhibit transitions between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), including snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}, amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

Predominant theories accounting for the spread of new sex-determining systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex ratio selection~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Tighter linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system.

The sex ratio is directly determined by the sex-determining system, and it has therefore been suggested that sex ratio selection is a dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
`Fisherian' sex ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the population sex ratio is biased towards one sex, the average per-individual contribution of genetic material to the next generation from the opposite sex is greater. 
Therefore, a mutant that increases investment in the rarer sex will spread via the higher per-individual contributions made by that sex. 
%The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determining system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex-determining systems. 
In the case of sex-chromosome evolution, Kozielska et al. (2014)~\cite{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased~\cite{Hamilton:1967ts}. 
They find that masculinizing or feminizing mutations that arise at a new, unlinked locus (i.e., neo-Y or neo-W chromosomes) can then spread, which restore an even sex ratio.

Here we use mathematical models to find the conditions under which new sex-determining systems spread when individuals experience selection at both diploid and haploid stages, which allows fitness differences between the sexes and sex ratio biases to occur simultaneously. 
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive~\cite{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (potentially from multiple parents); the term `haploid selection' encompasses both processes.

Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition~\cite{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, meiotic drive generally occurs either during the production of male or female gametes only~\cite{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex specific, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than randomly-chosen genes~\cite{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection (e.g.,~\cite{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}). 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain~\cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals can alter haploid allele frequencies and increase offspring fitness~\cite{Immler:2014im,Alavioon2017}.

There are various ways by which genes experiencing haploid selection could influence transitions between sex-determining systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in~\cite{vanDoorn:2007eu,vanDoorn:2010hu}). 
That is, new masculinizing mutations (neo-Y alleles) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
On the other hand, sex ratios can also become biased by linkage between the sex-determining locus and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles (Burt and Trivers, 2006, Chapter 6~\cite{Burt:2006}) or selection among X- and Y-bearing pollen~\cite{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex-determining systems would be influenced by the combination of sex ratio biases and associations with haploid selected alleles.

We find that the spread of novel sex-determining systems is influenced by both Fisherian sex ratio selection and by selection on linked alleles. 
%We find that sex ratio biases caused by haploid selection can exert Fisherian sex ratio selection upon novel sex-determiners but that their spread is also determined by selection on genetically-associated alleles. 
%Indeed, it is only when haploid-selected loci are tightly linked to the ancestral sex-determining region (and so sex ratio biases are initially large) that we see an asymmetry between selection for XY to ZW transitions and ZW to XY transitions (e.g., because haploid selection in males only causes biased zygotic sex ratios in an ancestrally XY system).
Surprisingly, Fisherian sex ratio selection does not dominate; it is possible for selection on linked alleles to drive turnover between sex-determining systems despite causing increasingly biased sex ratios. 
In addition to considering haploid selection, another novel development in our model is that we consider loci that are in very tight linkage with the ancestral sex-determining locus. 
%Because sex-determining loci are often found within a region of suppressed recombination, there can be a significant number of tightly linked loci. 
We find that loci tightly linked with the sex-determining locus can drive transitions in which the heterogametic sex changes, even when the new sex-determining locus is less closely linked to loci under selection (either including haploid selection or not).

\section*{Model}

%genetic system
We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles (Fig \ref{fig:model_outline}). 
Locus $\mathbf{X}$ is the ancestral sex-determining region, with alleles X and Y (or Z and W).
Locus $\mathbf{A}$ is a locus under selection, with alleles $A$ and $a$.
Locus $\mathbf{M}$ is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, $\mathbf{X}$; XX genotypes become females and XY become males (or ZW become females and ZZ become males). 
To evaluate the evolution of new sex-determining systems, we consider the invasion, fixation, maintenance, and/or loss of a novel sex-determining allele ($m$) at the $\mathbf{M}$ locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the $\mathbf{M}$ locus is epistatically dominant over the $\mathbf{X}$ locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the $\mathbf{X}$ locus genotype.
With $k=0$, the $m$ allele is a masculinizer (a neo-Y allele) and with $k=1$ the $m$ allele is a feminizer (a neo-W allele).
With intermediate $k$, we can interpret $m$ as an environmental sex-determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
%We also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Model Outline
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
%\centering
%\centerline{\includegraphics[width=\linewidth]{Sex_determination_outline.pdf}}
\caption{
{\bf Outline of model features.}
Panel A: Recombination rate parameters between the ancestral-sex-determining locus ($\mathbf{X}$, here assumed to have alleles X and Y), a locus under selection ($\mathbf{A}$, with alleles $A$ and $a$), and a new sex-determining locus ($\mathbf{M}$, with alleles $M$ and $m$). 
If $r<1/2$, then associations between ancestral sex-determining alleles and selected alleles can be maintained past recombination in males. 
Panel B: Haploid selection is often sex-specific, occurring during haploid production or competition in either males or females. 
For example, haploid selection in males only is represented by the dashed circle.  
If X or Y alleles remain associated with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X- or Y-bearing male gametes/gametophytes will be more abundant after haploid selection. 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determining systems. 
Similarly, zygotic sex ratio biases can occur if haploid selected alleles are associated with new sex-determining alleles ($R<1/2$). 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW), a neo-Y allele spreads to pseudo-fixation (its maximum frequency among male gametes) and the ancestral Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele spreads to pseudo-fixation (its maximum frequency among female gametes) and the ancestral X allele is lost. 
Neo-W alleles allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in \nameref{app:recurs}. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the $\mathbf{A}$ locus genotype; relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which implies a polygamous mating system. 
Gametic competition in monogamous mating systems is, however, equivalent to meiotic drive in our model (described below), as either only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, conjugation between male and female gametes occurs at random.
The resulting zygotes develop as males or females, depending on their genotypes at the $\mathbf{X}$ and $\mathbf{M}$ loci. %(and the $\mathbf{M}$ genotype of their mother in the case of maternal control).
Diploid males and females then experience selection, with relative fitnesses $w_{AA}^{\Hermaphrodite}$, $w_{Aa}^{\Hermaphrodite}$, and $w_{aa}^{\Hermaphrodite}$. %where $g$ is the diploid genotype at the $\mathbf{A}$ locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the $\mathbf{A}$ locus.
The next generation of gametes is produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci $\mathbf{X}$ and $\mathbf{A}$ with probability $r$, between loci $\mathbf{A}$ and $\mathbf{M}$ with probability $R$, and between loci $\mathbf{X}$ and $\mathbf{M}$ with probability $\rho$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\rho$ (see Fig \ref{fig:model_outline}A and \nameref{tab:chisubstitutions}). 
Individuals that are heterozygous at the $\mathbf{A}$ locus may experience meiotic drive; a gamete produced by $Aa$ heterozgotes of sex $\Hermaphrodite$ bears allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the $\mathbf{A}$ locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

\begin{table}[ht]
%\centering
\smallskip
\caption{{\bf Relative fitness of different genotypes in sex $\Hermaphrodite \in \{\female,\male\}$ } }
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  $A$ & $w_{A}^\Hermaphrodite = 1+t^\Hermaphrodite$ \\
  $a$ & $w_{a}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  $AA$ & $w_{AA}^\Hermaphrodite = 1+ s^\Hermaphrodite$ \\
  $Aa$ & $w_{Aa}^\Hermaphrodite = 1+h^\Hermaphrodite s^\Hermaphrodite$ \\
  $aa$ & $w_{aa}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Transmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  $A$ & $\alpha^\Hermaphrodite=1/2+\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  $a$ & $1-\alpha^\Hermaphrodite=1/2-\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%

The model outlined above describes both ancestral XY and ZW sex-determining systems. %if we relabel the two sexes as being ancestrally `heterogametic' or ancestrally `homogametic'. 
Without loss of generality, we refer to the ancestrally heterogametic sex as male and the ancestrally homogametic sex as female.
That is, we primarily describe an ancestral XY sex-determining system but our model is equally applicable to an ancestral ZW sex-determining system (relabelling the ancestrally heterogametic sex as female and the ancestrally homogametic sex as male and switching the labels of males and females throughout). 
We use a superscript to specify the ancestral sex-determining system described, e.g., $^{(XY)}$ for ancestral XY sex-determination.

\subsection*{Generic invasion by a neo-Y or neo-W}

The evolution of a new sex-determining system requires that a rare mutant allele, $m$, at the novel sex-determining locus, $\mathbf{M}$, increases in frequency when rare. 
This is determined by the leading eigenvalue ($\lambda_m^{(XY)}$) of the system of eight equations describing the frequency of eggs and sperm carrying the $m$ allele in the next generation (equations \ref{eq:recursions}). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially for a dominant neo-Y ($k=0$) or neo-W ($k=1$), see \nameref{app:inv_cond}. 
The leading eigenvalue for a rare neo-Y or neo-W allele, $m\in\{Y',W'\}$, is the largest value of X that solves $x^2+ b x + c = 0$.  %$(\lambda_m^{(i)})^2+ b \lambda_m^{(i)} + c =0$. 
The coefficients are $b= - (\Lambda_{mA}^{(XY)} + \Lambda_{ma}^{(XY)})+(\chi_{mA}^{(XY)} + \chi_{ma}^{(XY)})$ and $c = (\Lambda_{mA}^{(XY)} - \chi_{mA}^{(XY)}) (\Lambda_{ma}^{(XY)} - \chi_{ma}^{(XY)}) - \chi_{mA}^{(XY)} \chi_{ma}^{(XY)}$, where $\Lambda_{mi}^{(XY)}>0$ is the multiplicative growth rate (which we will call the ``haplotypic growth rate'') of the neo-sex determination allele $m$ on background $i\in\{A,a\}$ without accounting for loss due to recombination ($R=0$), and $\chi_{mi}^{(XY)}>0$ is the rate at which mutant haplotypes on background $i\in\{A,a\}$ recombine onto the other $\mathbf{A}$ locus background in heterozygotes (proportional to $R$), see Table \ref{tab:haplotype_growth}.
In the ancestral population, it is convenient to follow the frequency of the $A$ allele among female gametes (eggs), $p^\female_X$, and among X-bearing, $p^\male_X$, and among Y-bearing, $p^\male_Y$, male gametes (sperm/pollen). 
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males. 
We will consider only equilibrium frequencies of alleles, $\hat{p}^\Hermaphrodite_i$, and Y-bearing male gametes, $\hat{q}$, when calculating the eigenvalues.

\begin{table}[!ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{ {\bf Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   $m$ is a neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\Lambda_{Y'A}^{(XY)} = {\left( 2 \zeta \right)}^{-1} \left[\hat{p}_X^\female w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\Lambda_{Y'a}^{(XY)} = {\left( 2 \zeta \right)}^{-1} \left[(1-\hat{p}_X^\female) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha_{\Delta}^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Y'A}^{(XY)} = R {\left( 2 \zeta \right)}^{-1} \left[ (1-\hat{p}_X^\female) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} (1+\alpha_{\Delta}^{\male}) \right]/  \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)   $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{Y'a}^{(XY)} = R {\left( 2 \zeta \right)}^{-1} \left[   \hat{p}_X^\female w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha_\Delta^{\male}) \right]/ \left( \bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\male}_{D} \right)  $\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  $m$ is a neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\Lambda_{W'A}^{(XY)} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}(1+\alpha_{\Delta}^{\female})\right]/ \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\Lambda_{W'a}^{(XY)} = {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha_{\Delta}^{\female})\right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{W'A}^{(XY)} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} (1+\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right)  $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\chi_{W'a}^{(XY)} = R {\left[ 2 (1 - \zeta) \right]}^{-1} \left[ \bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha_{\Delta}^{\female}) \right] / \left(\bar{w}_H^\female \bar{w}_H^\male \bar{w}^{\female}_{D} \right) $ \\ [1ex]
  \hline \hline 
   \end{tabular}
   \begin{flushleft} 
$\bar{p}^{\male}=(1-\hat{q})\hat{p}_X^\male + \hat{q}\hat{p}_Y^\male$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
%      \item $R$ is the probability of recombination between loci $\mathbf{A}$ and $\mathbf{M}$.
$\zeta$ is the zygotic sex ratio (fraction male).
$\bar{w}^{\Hermaphrodite}_{D}$ is the mean fitness of diploids of sex $\Hermaphrodite$.
$\bar{w}_H^\Hermaphrodite$ is the mean fitness of haploids from sex $\Hermaphrodite$, see \nameref{tab:meanfitnesses}.
\end{flushleft}
  \label{tab:haplotype_growth}
\end{table}

The new sex-determining allele increases in frequency when rare when the largest eigenvalue is greater than one ($\lambda_m^{(XY)}>1$). 
%If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur, i.e., if $(\Lambda_{mA} + \Lambda_{ma})/2 > 1$ then $\Lambda > 1$. 
%The leading eigenvalue is the largest solution of $f(\lambda_m^{(i)})=(\lambda_m^{(i)})^2+b\lambda_m^{(i)}+c=0$ and the Perron-Frobenius theorem guarantees that it is positive, unique, and real.  
%Since $f(\Lambda_{mA}^{(i)})$ and $f(\Lambda_{ma}^{(i)})$ are of opposite signs, the leading eigenvalue must fall between these two quantities and is the larger of them when $R = 0$ (see \nameref{file:Mathematica}).  
If both haplotypic growth rates are greater than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}>1$), then the new sex-determining allele invades regardless the rate of recombination between the new sex-determining locus and the selected locus ($R$), see \nameref{app:inv_cond} for details.
Conversely, if both haplotypic growth rates are less than one ($\Lambda_{mA}^{(XY)}, \Lambda_{ma}^{(XY)}<1$), then invasion can never occur.  
Finally, if only one haplotypic growth rate is greater than one, the new sex-determining allele can always invade when arising at a locus that is tightly linked to the selected locus ($R\approx0$).  
Furthermore, it can be shown that the leading eigenvalue declines with $R$, and invasion requires that $R$ is sufficiently small such that: 
\begin{equation}\label{eq:lambdasGen}
%\frac{\chi_{mA}}{\Lambda_{mA}-1}+\frac{\chi_{ma}}{\Lambda_{ma}-1}>0.
\chi_{ma}^{(XY)}/\left(\Lambda_{ma}^{(XY)} - 1\right) + \chi_{mA}^{(XY)}/\left(\Lambda_{mA}^{(XY)} - 1\right) < 1.
\end{equation}
\noindent 
This condition may or may not be satisfied for the full range of locations of the new sex-determining locus, including $R=1/2$, depending on the nature of selection.  
Interpreting this condition, if we assume that only the $mA$ haplotype would increase in frequency when $R=0$ ($\Lambda_{ma}^{(XY)}<1<\Lambda_{mA}^{(XY)}$) the first term on the left-hand side of \eqref{eq:lambdasGen} is negative and invasion requires that rate of growth of $mA$ haplotypes ($\Lambda_{mA}^{(i)}-1>0$) and the rate at which $mA$ haplotypes are produced by recombination in $ma$ haplotypes ($\chi_{ma}^{(i)}$) are sufficiently large relative to the rate of decline of $ma$ haplotypes ($1-\Lambda_{ma}^{(i)}>0$) and the rate of loss of $mA$ haplotypes due to recombination ($\chi_{mA}^{(i)}$).

We can draw a number of key points about the invasion of neo-Y and neo-W mutations from Table \ref{tab:haplotype_growth}. 
First, Fisherian sex ratio selection will favour the spread of a neo-Y and inhibit the spread of a neo-W if the ancestral zygotic sex ratio is biased towards females, $\zeta<1/2$ (i.e., the first factor of the $\Lambda_{mi}^{(XY)}$ is greater than one for a neo-Y and less than one for a neo-W).
However, the spread of a neo-Y (neo-W) also depends on the male (female) fitness of associated alleles (see terms involving equilibrium allele frequencies, $\hat{p}$'s).
Second, invasion by a neo-Y (neo-W) does not directly depend on the fitness of female (male) diploids. %(for a given set of equilibrium allele frequencies).
This is because a dominant neo-Y (neo-W) is always found in males (females), and therefore the frequency of the neo-Y (neo-W), $m$, only changes in males (females), Fig \ref{fig:model_outline}C,D. 
Finally, invasions by a neo-Y and a neo-W are qualitatively different.
This is because a gamete with the neo-Y always pairs with a female gamete containing an X, Fig \ref{fig:model_outline}C.
By contrast, a gamete with a neo-W can pair with an X- or Y-bearing male gamete, Fig \ref{fig:model_outline}D.
Consequently, neo-W-bearing females obtain a different frequency of $A$ alleles from mating (when $\hat{p}_X^\male \neq \hat{p}_Y^\male$) compared to ancestral ($MM$) females.

>>>>

%reference equations in appendices
\usepackage{xr}
\externaldocument{S1Appendix}
\externaldocument{main}
\externaldocument{S3Appendix}

@@
